---
title: "Day 1 Part 1 - Data"
author: "Philip Parker"
date: "6 February 2015"
output:
  slidy_presentation:
     duration: 30
     css: css/reveal.css

---
## My first R program - Assignment
```{r}
# The symbol <- is used to assign something to an object.
myFirst <- "Hello World!"
myFirst
# Everything in R is a function. 
#Functions are called via a name then brackets
print(myFirst)
cat(myFirst)
#Lets do it a different way
myFirst2 <- c("Hello", "World", "!")
print(myFirst2)#yuck
paste(myFirst2, collapse = " ")# OK I suppose
cat(myFirst, sep = " ")# Nice
#We can assign the results of a function to an object
myFirst3 <- paste(myFirst2, collapse = " ")
myFirst3
#Be Carful as you can easily overwrite an object
myFirst3 <- "Hello Joe"
myFirst3# oh no!!!!
```


## R basics (the boring bits) - Data forms
Forms data can take include:

- Integer: 1, 2, 3
- Numeric: 1.0, .2, 3.6
- Logical: TRUE, FALSE
- Character: "one", "two", "three"
- Missing: NA

## Vectors
A vector is the simpliest data type. All elements have to have the same form
```{r}
#vector - Interger
c(1,2,3,4, NA)
# Numeric
c(1.3,4.0,5,NA,7.5, .2)
#Logical
c(NA,TRUE,FALSE,TRUE)
#Character
c("One", "Fish", "Two", "Fish", NA)
#This will not work - will convert everything to a string
c(1,TRUE,"one")
```

## Factors 
Factors are special character vectors
```{r}
#Nominal
gender <- c(rep("male",5), rep("female", 10)) 
gender <- factor(gender) 
gender
#Ordinal
height <- c(rep("Phil",5), rep("Joe", 5), rep("Herb", 5))
height <- factor(height, ordered=TRUE)
height
```


## Lists
A list is a collection of objects of any type or form
```{r}
int <- c(1,2,3,4, NA)
num <- 3.14
log <- "FALSE"
char <- c("One", "Fish", "Two", "Fish", NA)

myList <- list(int=int,num=num,log=log,char=char, phrase="Hello World")
myList
```

## Matrix
A matrix is a two dimensional vector (rows, columns) of the same data form
```{r}
myMatrix <- matrix(1:100, nrow=10, ncol=10)
myMatrix
#not So helpful
colnames(myMatrix) <- paste0("C_",letters[1:10])
row.names(myMatrix)<- paste0("R_", letters[1:10])
myMatrix
```

##Data Frame
A data.frame is a structured list (rows, columns) of any data form. This is the ONLY format that SPSS uses.

```{r}
myData <- data.frame(var1 = 1:10, var2 = letters[1:10], var3=rep(c(TRUE,FALSE), 5))
myData
```

This format is mostly how you want your data read into R

## Data
Suprisingly most people give up on R in the first few minutes because they cannot get there data into R. Here are the main things you need:

- Know the path to where your data is located
- Know the working directory R is using
- Know the format that your data is in
- Know how to check your data is there
- Don't use the point and click features! 

## Data - where is it

- R is based on unix OS so using "/" to seperate folders rather than "\". 
- Your default home location will be represented by a "~" sign  
- Lets see if a file exists (use the 'tab' key for code completion)
```{r}
file.exists("~/Dropbox/Rcourse/data/PISA2003.sav")
```
- Getting and setting Rs working directory
```{r}
getwd()
setwd("~/Dropbox/Rcourse/data")
file.exists("PISA2003.sav")

```
- Read in the data
```{r, eval=FALSE}
#Assumes you have set the working directory!!
#csv data
myData <- read.csv("myCSV.csv", header=TRUE, na.strings = c(-97, -98, -999) )
#txt file - tab deliminated
myData <- read.table("myText.txt", header = TRUE, sep = "\t", na.strings = -999)
#Read in SPSS data
library(foreign)
myData <- read.spss("mySPSS.sav", to.data.frame = TRUE, use.value.labels = FALSE)
```

- Foreign will also read in minitab, SAS, stata, etc.
- R also has great support for SQL and other databases.


