---
title: "Day 1 Part 2 - Functions"
author: "Philip Parker"
date: "6 February 2015"
output:
  slidy_presentation:
     duration: 30
     css: css/reveal.css

---


## Writing your own functions

- The general rule in R is if you find yourself doing something repeatedly make it a variable or a function
- One of the major benefits of R is the ability to write your own functions
- It is actually pretty easy to do. Here is a 'Hello World' function

```{r}
myFun <- function(){print("Hello World!!")} 
myFun()
```

## Writing your own functions: The details

The recipe for a function is as follows:

- An object you want to assign your function to 'myFun <-'
- The command function followed by arguments you want to supply 'function()'
- The code you want the function to run and what you want it to return in curly brackets
- Lets extend the function above to make it more personal

## Writing your own functions: The details

```{r}
#function with 1 argument
myFun <- function(yourName){
  temp <- paste("Hello", yourName, "!!!" )
  cat(temp)
  }

myFun("Phil")
# function with 1 argument with a default setting
myFun <- function(yourName = "Phil"){
  temp <- paste("Hello", yourName, "!!!" )
  cat(temp)
  }

myFun()
myFun("Bob")
```

## Make it crazy
```{r, eval=FALSE}
#You can make it as crazy as you want
myFun <- function(yourName = "Phil"){
  temp <- paste("Hello", yourName, sep = "+")
  url <- paste0("http://tts-api.com/tts.mp3?q=", temp, ".")
  browseURL(url)
  }

myFun()
 myFun("Bob")
```

## Multiple argument functions

- Often I want both the central tendency and some measure of spread
- But maybe I want it to be flexible enough to handle either parametric or non-parametric

```{r}
mySummary <- function(x, parametric = "both"){
  #x = a vector of type numeric or integer
  #parametric = should function return parametric ('par'), non-parametric ('npar'), or both ('both')
  if (parametric == 'both'){
    temp = c(Mean = mean(x, na.rm=TRUE), SD = sd(x, na.rm=TRUE), 
             Median = median(x, na.rm=TRUE), MAD = mad(x, na.rm=TRUE) )
  }else if (parametric == 'par'){
    temp = c(Mean = mean(x, na.rm=TRUE), SD = sd(x, na.rm=TRUE) )
  }else {
    temp = c(Median = median(x, na.rm=TRUE), MAD = mad(x, na.rm=TRUE) )
  }
  temp
}

mySummary(c(1,2,3,4,5,6,7))
```

