---
title: "Day 1 Part 2 - Playing with Data"
author: "Philip Parker"
date: "6 February 2015"
output:
  slidy_presentation:
     duration: 30
     css: css/reveal.css

---


## Basic Descriptives
 - For the rest of the day we will work with a single clean dataset with some missing.
 - We are also going to add new column: id, schoolid, gender
 - To do this we will have to take the date from class matrix to class data.frame
```{r, results='hide'}
library(psych)
data(ability)
class(ability)
ability <- as.data.frame(ability)
names(ability)
#add new variable
ability$id <- 1:nrow(ability)
ability$school.id <- rep(1:61, each = 25)
ability$gender <- "F"
ability[sample(1:1525, 600), "gender"] <- "M"
ability$gender <- factor(ability$gender)
names(ability)
str(ability)
```

## Basic Descriptives - Single Variable
```{r}
# This will strangely return NA
mean(ability$reason.4)
# Default action for missing in R is to fail. Need to tell it don't do that
mean(ability$reason.4, na.rm=TRUE)
```

- Try median, range, min, max, table
- Read the help menu on quantile and return ONLY the 2.5 and 97.5% quantiles

## Basic Descriptives - Multiple Variables
```{r, eval=FALSE}
summary(ability)
describe(ability)
```
- Read documentation on describeBy and get statistics for the FIRST 4 variables by gender

## Apply family
- The apply family is R's compact version of for loops.
- We will look at apply and tapply
```{r}
apply(ability[,1:4], 2, mean, na.rm=TRUE)
tapply(ability$reason.4, ability$gender, mean, na.rm=TRUE)
#scale score
ability$reason <- apply(ability[,1:4],1,mean, na.rm=TRUE)
```

- Create scale scores for the other variables 
- Use functions learnt above to explore the new variables
- Plot the new variables: Try plot, hist, boxplot
- We will look at better ways of creating scale scores next time

## Subsetting and splitting
```{r}
# Splitting data
schoolAbility <- split(ability, ability$school.id)
# Subsetting data
girls <- subset(ability, subset = gender == "F")
boys <- subset(ability, subset = gender == "M")
```

- lookup the lapply and sapply commands. Can you get summary statistics for each school?

## Merge
There are three main merging comands in r:

- rbind or rbind.data.frame for stacking data (rbind.fill in the plyr package for unbalanced data)
- cbind or cbind.data.frame for putting columns together
- merge
```{r, eval=FALSE}
#Stack data
ability2 <- rbind.data.frame(girls,boys)
#Create some new data to add
myMatrix <- matrix(rnorm(1525*5, 5,2), nrow=1525)
colnames(myMatrix) <- c("MathSC", "EngSC", "SocialSC", "GeneralSC", "ParentSC")
# Merge using cbind
ability3 <- cbind.data.frame(ability, myMatrix)
#Merge by ID
myPersonality <- data.frame(myMatrix)
myPersonality$id <- sample(1:1525, 1525)
ability4 <- merge(ability, myPersonality, by = "id")
```

## Aggregation
- Ok so no IPPE presentation is complete without reference to the BFLP
- So how would we add a school average reason score to our data
- We need both aggregate, name, and merge functions
```{r}
schoolAvg <- aggregate(ability$reason, by = list(ability$school.id), mean, na.rm=TRUE)
names(schoolAvg) <- c("school.id", "avg.reason")
abilityBFLPE <- merge(ability, schoolAvg, by = "school.id")
```
- How might you aggregate multple variables at a time
- How could we aggregate by both school.id and gender

