---
title: "Day 1 Part 1"
author: "Philip Parker"
date: "6 February 2015"
output:
  slidy_presentation:
     duration: 30
     css: css/reveal.css

---

## Why R

- Free
- Comprehensive
- Flexible
- Does everything (interactive websites, reproducible research)
- In some cases you have no choice
- Great support for [parallel](http://www.r-bloggers.com/how-to-go-parallel-in-r-basics-tips/) processing
- It's fast...er
- Plenty of support from multiple data sources (e.g., Facebook, Twitter, weather, census data, web scrapping)
- Established procedures for downloading, storing, managing, and analysing many large survey data (e.g. PISA)
- R integrates and extends mplus, spss, python, C, Fortran, SQL, HTML/Markdown, ...


## Why not R
- It is a stupid inconsistent [language](http://www.burns-stat.com/pages/Tutor/R_inferno.pdf)
- It is a full language not a macro language (e.g. spss). The learning curve is steep!!
- It is community supported so no one is paid to be nice to you
- Error messages are designed for those familiar with the language


## Some things are just easier in R
Some things are just easier in R:

- Say goodbye to reverse scored items
- Dummy coding
- Multilevel modelling
- glm
- SEM

## Did I mention pretty graphs?
```{r, echo=FALSE}
# Earth at night in R
# took about 30 minutes to code

library(maps)
data(world.cities)
#subset minor and major cities
minmajdat=world.cities[world.cities$pop>10000,]

#subset major cities
majdat=world.cities[world.cities$pop>100000,]

#subset huge cities
hdat=world.cities[world.cities$pop>1000000,]

ranker=rank(world.cities$pop)

par(bg=rgb(0,0,0.025),mex=0.5,mar=c(0,0,0,0))

map("world",col=rgb(0,0,0.55,
alpha=0.05),interior=F,fill=T)

#plot huge cities
points(hdat$lat[ranker]~hdat$long[ranker],col=rgb(0.25,0.25,0,alpha=0.05)
,pch='.',cex=(hdat$pop)^c(1/100))

# illumate the night

  for(i in 1:5){
points(jitter(world.cities$lat[ranker],1)~jitter(world.cities$long[ranker],1),
col=rgb(230/255,232/255, 250/255,alpha=0.0075),pch='.',cex=(world.cities$pop)^c(1/100))
	}

#create light haze arround minor & major cities
	for(i in 1:5){
icol=rgb(0.7,0.7,0.7,alpha=0.02)
points(jitter(minmajdat$lat[ranker],10)~
jitter(minmajdat$long[ranker],10),
col=icol,pch='.',cex=(minmajdat$pop)^c(1/100))
	}

#create yellow light haze arround major cities
	for(i in 1:6){
icol=rgb(1,1,0,alpha=0.02)
points(jitter(majdat$lat[ranker],10)~
jitter(majdat$long[ranker],10),
col=icol,pch='.',cex=(majdat$pop)^c(1/100))
	}

#create bright light haze arround huge cities
	for(i in 1:6){
icol=rgb(1,1,1,alpha=0.02)
points(jitter(hdat$lat[ranker],10)~
jitter(hdat$long[ranker],10),
col=icol,pch='.',cex=(hdat$pop)^c(1/100))
	}
```

## Interactive graphs
<iframe src="http://rpubs.com/PhilipParker/plots2" width="800" height="800"></iframe>


## Pretend you are working!
```{r, eval=FALSE}
## install.packages('fun')
library(fun)
mine_sweeper()
gomoku()
```


## Installing R
Notes for how to install are located [here](https://github.com/pdparker/rcourse)

- R is the programming language and can be installed [here](http://cran.r-project.org/)
- R studio is the better GUI and can be installed [here](http://www.rstudio.com/products/rstudio/download/)

Rstudio now covers:

  - R programming
  - R graphics
  - R markdown for word documents, PDFs, web pages
  - Shiny for interactive web pages
  - Git for version control
  - We will only work with R programming in this course
  
Lets look at Rstudio now
  
## Reproducible Research
Rs greatest strength is reproducibility. To leverage this:

- Don't be lazy
- Comment comment comment
- Markdown is a super super easy language integrated into Rstudio for html, word, pdf, or slides
- Go do a git course. Absolutely life saving!

```{r, eval=FALSE}
#This is a comment. Have lots of them in your script.
```

## Reproducible Research
- Never use point and click menus (only time saving the FIRST time!!)
- Have a quarantined data file. OR just use SQL and only give collaborators read permission 
- Only create new data files if you have to!
- Make your code clean, use indentation.
- Use descriptive names - there is no silly 8 character limits for variable names in R
- BE CONSISTENT!!! I use camelCase for everything


