---
title: "Day 1 Part 1 - Data"
author: "Philip Parker"
date: "6 February 2015"
output:
  slidy_presentation:
     duration: 30
     css: css/reveal.css

---
## R basics (the boring bits) - Introduction to Object Orientated Programming

This is NOT a technical introduction but a pragmatic introduction.

Everything in R is:

- An element of a given type
- An object (think container)
- A function

## R basics (the boring bits) - Elements
'Types' data can take include:

- Integer: 1, 2, 3
- Numeric: 1.0, .2, 3.6
- Logical: TRUE, FALSE
- Character: "one", "two", "three"
- Missing or otherwise indisposed: NA, NULL, NaN

## Objects
Objects in R are just containers that you put elements into for storage.

To put something in a container we use the assignment symbols:

- objectName <- stuff you want put in the container

Containers come in the form of:

 - Vectors: 1 or more elements of the same type e.g. 1,2,3
 - Factors: character vector with 'special' properties
 - List: 1 or more objects of any type 1,"A", TRUE
 - Matrix: A two-dimensional vector with attributes [rows:columns]
 - Dataframe: A two-dimensional list with attributes [rows:columns]
 - Array: an N dimensional vector
 
## Functions

A function takes input, transforms it, and returns output. Examples include:

- mean, sd, range, min, max
- lm, t.test, aov, anova, sem, cfa
- summary, plot, str, names, print, cat

## My first R program - Assignment {.smaller}
We are going to take a vector of characters assign to an object and apply a function to it
```{r}
# Object <- vector of character elements
myFirst <- c("Hello","World!")
#Function cat returning the elements in a nice format
cat(myFirst, sep = " ")
# Lets assign the output of a function to an object
mySecond <- paste(myFirst, collapse = " ")
# We can now apply functions to this and functions within functions. 
# Lets shout out excitement at learning R
cat(toupper(mySecond), sep = " ")
# Another way to do this you might find easier to follow
library(magrittr)
myFirst %<>% paste(collapse = " ") %>% toupper %>% cat
```

##Other functions to play with
- Beginner: nchar, tolower
- Intermediate: strsplit, substring
- Advanced: gsub, plot with text


## Vectors
A vector is the simplest data type. All elements have to have the same form
```{r}
#vector - Integer
c(1,2,3,4, NA)
# Numeric
c(1.3,4.0,5,NA,7.5, .2)
#Logical
c(NA,TRUE,FALSE,TRUE)
#Character
c("One", "Fish", "Two", "Fish", NA)
#This will not work - will convert everything to a string
c(1,TRUE,"one")
```

## Vector Indexing
```{r}
# Example vector
myVect <- c(1.3,4.0,5,NA,7.5, .2)
# Extract element 3
myVect[3]
# Extract elements 3 to 5
myVect[3:5]
# Extract non-consecutive elements
myVect[c(1,3,6)]
# Which elemnts are greater than 3
which(myVect > 3) # can also use ==, <, <=, >=, !=
#Extract them
myVect[which(myVect > 3)]
```

## Factors 
Factors are special character vectors
```{r}
#Nominal
gender <- c(rep("male",5), rep("female", 10)) 
gender <- factor(gender) 
gender
#Ordinal
height <- c(rep("Phil",5), rep("Joe", 5), rep("Herb", 5))
height <- factor(height, ordered=TRUE)
height
```

## Functions to try
- table and chisq.test
- relevel

## Lists
A list is a collection of objects of any type or form
```{r}
int <- c(1,2,3,4, NA)
num <- 3.14
log <- "FALSE"
char <- c("One", "Fish", "Two", "Fish", NA)

myList <- list(int=int,num=num,log=log,char=char, phrase="Hello World")
myList
```

## Indexing lists
```{r}
#By name
myList[["char"]]
# By number
myList[[4]]
```


## Matrix
A matrix is a two dimensional vector (rows, columns) of the same data form
```{r}
myMatrix <- matrix(1:100, nrow=10, ncol=10)
myMatrix
#not So helpful
colnames(myMatrix) <- paste0("C_",letters[1:10])
row.names(myMatrix)<- paste0("R_", letters[1:10])
myMatrix
```

##Indexing Matrices
- Has the form [row,column]
- Leaving one side empty is short hand for "give me all of them""
```{r}
# All columns and row 3
myMatrix[3,]
# All rows and column 4
myMatrix[,4]
# Cell row 3 column 3
myMatrix[3,3]
```
- Try diag, upper.tri, lower.tri
- As an example change the upper triangular elements to 0

## Data Frame
A data.frame is a structured list (rows, columns) of any data form. This is the ONLY format that SPSS uses.

```{r}
myData <- data.frame(var1 = 1:10, var2 = letters[1:10], var3=rep(c(TRUE,FALSE), 5))
head(myData)
names(myData)
str(myData)
```

## Indexing Data Frames
```{r}
#By name
myData$var1
#Using matrix indexing
myData[1,1:2]
# By list indexing
myData[["var1"]]
```

## Data
Surprisingly most people give up on R in the first few minutes because they cannot get there data into R. Here are the main things you need:

- Know the path to where your data is located
- Know the working directory R is using
- Know the format that your data is in
- Know how to check your data is there
- Don't use the point and click features! 

## Data - where is it

- R is based on unix OS so using "/" to separate folders rather than "\\". 
- Your default home location will be represented by a "~" sign  
- Lets see if a file exists (use the 'tab' key for code completion)
```{r}
file.exists("~/Dropbox/Rcourse/data/PISA2003.sav")
```
- Getting and setting Rs working directory
```{r}
getwd()
setwd("~/Dropbox/Rcourse/data")
file.exists("PISA2003.sav")

```
- Read in the data
```{r, eval=FALSE}
#Assumes you have set the working directory!!
#csv data
myData <- read.csv("myCSV.csv", header=TRUE, na.strings = c(-97, -98, -999) )
#txt file - tab deliminated
myData <- read.table("myText.txt", header = TRUE, sep = "\t", na.strings = -999)
#Read in SPSS data
library(foreign)
myData <- read.spss("mySPSS.sav", to.data.frame = TRUE, use.value.labels = FALSE)
```

- Foreign will also read in minitab, SAS, stata, etc.
- R also has great support for SQL and other databases.

## Installing and using packages
```{r, eval=FALSE}
install.packages("MplusAutomation")
library(MplusAutomation)
help(MplusAutomation)
?prepareMplusData
```

## Exporting data
- csv: ?write.csv
- txt: ?write.table
- spss: ?write.foreign
- Mplus: ?prepareMplusData

## Pulling it all together
```{r}
#Set the working directory
setwd("~/Dropbox/Rcourse/data")
#Load the required packages
library(foreign)# read spss data
library(MplusAutomation) # write for spss
#read the data
myData <- read.spss("PISA2003-Achievement.sav", to.data.frame = TRUE, use.value.labels = FALSE)
#Check it read in ok
dim(myData)
names(myData)
#Advanced! Make that easier to read
sprintf("The number of cases is: %d and the number of variables is: %d", 
        nrow(myData), length(myData)
        )
```

## Writing to Mplus
```{r}
#Write to mplus file and get a mplus template file
prepareMplusData(myData, filename = "mplus.dat")
#Advanced! Write the template file to a folder
#Here I have put a dummy file location so you should 
capture.output(prepareMplusData(myData, filename = "mplus.dat"), file = "mplus.inp")
```


