---
title: "Day 1 Part 1 - Data"
author: "Philip Parker"
date: "6 February 2015"
output:
  slidy_presentation:
     duration: 30
     css: css/reveal.css

---


## Example data
- R has many example data sets so does most packages. To see a list of them
```{r, eval=FALSE}
data()
```
- To load an example data set
```{r}
data(iris)
head(iris)
```

## Toy data
- R is great for simulation
- Maybe we want to create a toy dataset to play with

## Basics of simulation
- Bias = The expected value is not the population value
- Consistency = How often is the observed value equal to the expected value
- Efficiency = How tightly packed are repeated observations

## Simulation - Bias
Population has a $\mu = 0$ and $\sigma = 1$
```{r}
myFunction <- function(x) sprintf("The mean is %.2f and the SD is %.2f", mean(x), sd(x))
#bias
myFunction(replicate(100, mean(rchisq(100,2))))
```

## Consistency
```{r}
#consistency
# n = 1
sum(abs(replicate(100, mean(rnorm(1)))) < .5)
# n = 10 
sum(abs(replicate(100, mean(rnorm(10)))) < .5)
```
## Efficency
```{r}
# Efficency
#n = 1
sd(replicate(100, mean(rnorm(1))))
# n = 10
sd(replicate(100, mean(rnorm(10))))
# n = 10000
sd(replicate(100, mean(rnorm(10000))))
```

## Regression1
A regression equation is:
$Y = \alpha + \beta_{1} X + \epsilon$

- Y is a continuous response variable
- alpha is mean value conditioned on the other predictors
- beta is weight applied to the predictor X
- epsilon is residual variance distributed as $N \sim (0,\sigma)$

## Regression2
```{r}
n <- 100
x <- rnorm(n, mean = 10, sd = 5)
y.hat <- 2 + 1.2*x
y <- y.hat + rnorm(n, mean = 0, sd = 3)
myD <- data.frame(x = x, y.observed = y)
model1 <- lm(y ~ x, data = myD)
summary(model1)
```

## Regression3
I ran this with n = 10,100,1000, 10 000 the results where:

| n         | alpha | beta |
|-----------|-------|------|
| simulated | 2     | 1.2  |
| 10        | .322  | 1.37 |
| 100       | 2.43  | 1.16 |
| 1000      | 1.85  | 1.22 |
| 10,000    | 2.07  | 1.19 |

## Simulation from Correlation or Covariance matrix
```{r}
library(MASS)
Sigma <- matrix(c(1,.8082,.8082,1),2,2)
Sigma
myD <- mvrnorm(n=1000, rep(0, 2), Sigma)
cor(myD)
```

## Getting help
In R:

- Error messages - READ THEM!
- ?functionName

Help files have the following structure:

- Usage: function(argument = 1, argumentYouSupply, ...)
- Arguments: Explanations of what the arguments in useage are for and do
- Sometimes there is a list of what the function returns (useful!)
- Examples of how to use the function (USE THESE!!!)

## Outside of R you need to ask good questions

```{r}
library(fortunes)
fortune("Brain surgery")
```

## Outside of R you need to ask good questions
- Do your homework first and debug  - also read the error message you are getting!
- Reduced your problem to the simplest form that still returns an error (usually isolates the problem)
- Google
- Use meaningful titles (I see 5 to 10 "Please Help" titled threads a day!)
- Be specific, be direct, explain in full, re-read to check that the question would make sense to someone other than you
- Describe what you are trying to do not the outcome
- Give example data and code tried (this is what simulation and toy data are for)
- Don't just give a function but give the package it came from
