---
title: "Day 1 Part 2 - Playing with Data"
author: "Philip Parker"
date: "6 February 2015"
output:
  slidy_presentation:
     duration: 30
     css: css/reveal.css

---


## Basic Descriptives
 - For the rest of the day we will work with two subsamples of the PISA2003 data
 - Lets read in the data first 
```{r, results='hide', warning=FALSE,message=FALSE}
setwd("~/Dropbox/Rcourse/data/")
library(foreign)
pisaPsych <- read.spss("PISA2003.sav", to.data.frame=TRUE, use.value.labels = FALSE)
pisaAch <- read.spss("PISA2003-Achievement.sav", to.data.frame=TRUE, use.value.labels = FALSE)
names(pisaPsych)
str(pisaPsych)
```

## Basic Descriptives - Single Variable
```{r}
# This will strangely return NA
mean(pisaPsych$HISEI)
# Default action for missing in R is to fail. Need to tell it don't do that
mean(pisaPsych$HISEI, na.rm=TRUE)
```

- Try median, range, min, max, table
- Read the help menu on quantile and return ONLY the 2.5 and 97.5% quantiles

## Basic Descriptives - Multiple Variables
```{r, eval=FALSE}
summary(pisaPsych)
# A better set of summary data is in psych package
library(psych)
describe(pisaPsych)
```
- Read documentation on describeBy and get statistics for the FIRST 4 variables by gender

## Apply family
- The apply family is R's compact version of for loops.
- We will look at apply and tapply
- Lets get the mean values for the achievement data
```{r}
apply(pisaAch[,c("PV1MATH", "PV1READ", "PV1SCIE", "PV1PROB")], 2, mean, na.rm=TRUE)
tapply(pisaAch$PV1PROB, pisaAch$SCHOOLID, mean, na.rm=TRUE)
# Average achievement across the four domains NOTE use of 1 instead of 2
pisaAch$GPA <- apply(pisaAch[,c("PV1MATH", "PV1READ", "PV1SCIE", "PV1PROB")],1,mean, na.rm=TRUE)
```

- Create scale scores for the other variables 
- Use functions learnt above to explore the new variables
- Plot the new variables: Try plot, hist, boxplot
- We will look at better ways of creating scale scores next time

## Subsetting and splitting
```{r}
# Splitting data
schoolAbility <- split(pisaAch, pisaAch$SCHOOLID)
# Subsetting data
girls <- subset(pisaPsych, subset = gender == 1)
boys <- subset(pisaPsych, subset = gender == 2)
```

- lookup the lapply and sapply commands. Can you get summary statistics for each school?

## Merge
There are three main merging comands in r:

- rbind or rbind.data.frame for stacking data (rbind.fill in the plyr package for unbalanced data)
- cbind or cbind.data.frame for putting columns together
- merge
```{r}
#Stack data
pisaPsych <- rbind.data.frame(girls,boys)
#Merge by ID
pisaAllData <- merge(pisaPsych, pisaAch, by = "STIDSTD")
#check
sprintf("The number of CASE in pisaPsych is %d, in pisaAch is %d, and in the merged data is %d/n", 
        nrow(pisaPsych),nrow(pisaAch), nrow(pisaAllData)
        )
sprintf("The number of VARIABLES in pisaPsych is %d, in pisaAch is %d, and in the merged data is %d", 
        length(pisaPsych),length(pisaAch), length(pisaAllData)
        )
```

## Aggregation
- Ok so no IPPE presentation is complete without reference to the BFLP
- So how would we add a school average reason score to our data
- We need both aggregate, name, and merge functions
```{r}
schoolAvg <- aggregate(pisaAllData$GPA, by = list(pisaAllData$SCHOOLID.x), mean, na.rm=TRUE)
names(schoolAvg) <- c("SCHOOLID.x", "avg.reason")
pisaAllData <- merge(pisaAllData, schoolAvg, by = "SCHOOLID.x")
```
- How might you aggregate multple variables at a time
- How could we aggregate by both school.id and gender

