---
title: "Exercises - Part1"
author: "Philip Parker"
date: "18 February 2015"
output: html_document
---

Simulation Regression
--------------

1. Create and save a new R script
2. Use the following template to answer the questions

Questions:
-------------

1. At what size n do the results start to converge to the population values?
2. What size sample do you need before both parameters are reliably significant? 
3. How could you introduce bias into the intercept and/or the regression slope?
4. Using the rnorm command how could you simulate a t.test or even anova?
5. The model1 and model1.sum objects. Have a look inside them!

```{r, eval=FALSE}
# The number of participants
n <- 1000
# A normally distributed predictor variable
x <- rnorm(n, mean = 10, sd = 5)
# The predicted response variable
y.hat <- 2 + 1.2*x
# The observed response variable
y <- y.hat + rnorm(n, mean = 0, sd = 3)
#Wrap observed response and predictor variables into a data.frame
myD <- data.frame(x = x, y.observed = y)
# Run a simple regression
model1 <- lm(y ~ x, data = myD)
# Extract the summary statistics
model1.sum <- summary(model1)
#Display results
model1.sum
##END##  
```

Simulation from a correlation matrix
----------

1. Create and save a new R script
2. Use the following template to answer the questions

Questions:
-------------
Preliminary:

1. Extract the correlation in row 1 column 2 from sigma
2. Extract ONLY the lower triangular items from sigma
3. Extract elements in sigma that are greater than .4

Simulation data from correlation matrix:

1. What is the relationship between origin social class, education, and destination social class?
2. How does this change as n changes?
3. Using lm as above run a regression model with destination social class predicted by origin social class. 
4. Do so again but include education?
5. In the pdf BD1967 p.63 a set of correlations are presented for Wisconsin. Generate data from these correlations are compare results for this data with the Blau and Duncan model.

```{r}
#install package MASS
# note you only need to run this ONCE!!

## UNCOMMENT LINE BELOW TO RUN ##
#install.packages("MASS")

# load package
library(MASS)
# Correlation matrix from Blau and Duncan (1967)
sigma <- matrix(c(1,.538,.374,
                  .538,1,.445,
                  .374,.445,1), 3,3)
# Colnames for variables
colnames(sigma) <- c("destSocialClass", "Education", "originSocialClass")
# Row names for variables
row.names(sigma) <- colnames(sigma)
# Simulate data
  # n = number of participants
  # mu = mean variables
  # Sigma = correlation or covariance matrix
BD1967Data <- mvrnorm(n = 200, mu = rep(0,3), Sigma = sigma)
```


